A general model for the evolution of nuptial gift-giving

### 2024-12-20 ms.Rmd
Update last comments based on round of revisions and Evol Lett comments

---
title: "A general model for the evolution of nuptial gift-giving"
author: |
  | Author list (anonymised)
date: '[1] Author institutions and email'
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
  html_document: default
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - evolution-letters.csl
    reference_docx: docx_template.docx
bibliography: nuptial_refs.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{natbib}
- \usepackage{lineno}
- \usepackage{caption}
- \usepackage[utf8]{inputenc}
- \bibliographystyle{amnatnat}
- \linenumbers
- \modulolinenumbers[2]
linestretch: 2
link-citations: yes
linkcolor: blue
csl: evolution-letters.csl
biblio-style: apalike
---

@@

We ran simulations in which male search time and female choosiness evolved from an ancestral state of no searching and no choosiness. 
In all simulations, $N$ was initialised at 1000 and $K = 1000$. 
Simulations ran for $t_{max} = 10000$ time steps. 
We set $T_{\mathrm{f}} = 2$, $\psi = 3$, and $\lambda = 1$ for all simulations, and we simulated across a range of $\alpha = \{0.1, 0.2, ..., 1.9, 2.0\}$ and $\gamma = \{0, 0.1, ..., 1.9, 2.0\}$ parameter values for 5000 replicates.
In a sensitivity analysis, we also re-ran simulations for values of $T_{\mathrm{f}} = 10$ and $\psi = \{1, 2, 4, 6 \}$ (see Supporting Information S3), and simulated the evolution of male nuptial gift searching and female choosiness (fixing $T_{\mathrm{m}} = \alpha$) separately (see Supporting Information S4).
Summary statistics for mean trait values, population size, sex ratios, proportion of females and males in and out of the mating pool, and mean number of encounters per female and male within the mating pool were all calculated in the last time step. 
The C code used for simulating these IBMs also allows for the reporting of summary statistics in each time step. 
Additionally, it can simulate explicit space and individual movement through the landscape. 
A neutral evolving trait was also modelled to ensure that the code functioned as intended, and processes were compartmentalised into individual functions to facilitate code testing. 
All code is publicly available on GitHub (https://github.com/AUTHOR_NAME/REPOSITORY_NAME).

<!---

A set of simulations with a value of $\gamma$ calculated from empirical data was also conducted (100 replicates). 
Here, $\gamma$ was parameterised using data on egg production as a function of eating nuptial gifts (see below). 
Additional simulation sets with lower and upper bounds of the estimated $\gamma$ were subsequently run. 
In the simulation sets with experimentally derived parameter values, all other parameters were identical to previous simulation batches.

--->

@@

For males, if nuptial gifts are not abundant and thus require a long time to find (i.e., high $\alpha$), or if males encounter many females per unit time (i.e., high $R / (1+\beta)$), then the nuptial gift must result in a high fitness increment for selection to favour searching. 
For females, if $\gamma$ is sufficiently high, fitness is increased by rejecting males without gifts and mating only with males that provide nuptial gifts. 
As offspring processing time ($T_{\mathrm{f}}$), mate encounter rate ($R\beta / (\beta + 1)$), or the probability of a male finding a nuptial gift ($1 - \exp(-T_{\mathrm{m}}/\alpha)$ ) decrease, the threshold value of fitness above which selection will favour choosiness ($\gamma_{\mathrm{f}}$) increases.
This can be understood intuitively by realising that rejecting a prospective male represents an opportunity cost for the female.

```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.cap = "The coevolution of male search and female choosiness as a function of nuptial gift search time ($\\alpha$). Points show where the lower 95% confidence interval of female choosiness (red) and male search (blue) exceeds zero, indicating evolution of choosiness or nuptial gift search. Each point includes data from 5000 replicate simulations with identical starting conditions. Zones are identified that correspond to areas where males search and females are choosy (D), males search but females are not choosy (B), and males do not search and females are not choosy (A), as also depicted in Figure 2. The number of individuals in the population remained at or near carrying capacity of $K = 1000$. In each time step, up to 3000 total pair-wise interactions occurred. Expected female processing time was set to $T_{\\mathrm{f}}=2$ time steps, and $\\gamma$ and $\\alpha$ values in the range [0.0, 2.0] and [0.1, 2.0], respectively, were used."}
SUMM         <- read.csv(file = "inst/summary_coevolving.csv");
alpha_vals   <- sort(unique(SUMM[,1]));  
TM_vals      <- tapply(X = SUMM[,3],  INDEX = SUMM[,1], FUN = mean);
m_vals       <- tapply(X = SUMM[,13], INDEX = SUMM[,1], FUN = mean);
beta_vals    <- tapply(X = SUMM[,14], INDEX = SUMM[,1], FUN = mean);
Male_M       <- tapply(X = SUMM[,15], INDEX = SUMM[,1], FUN = mean);
Female_M     <- tapply(X = SUMM[,16], INDEX = SUMM[,1], FUN = mean);
Mf_nupt      <- tapply(X = SUMM[,17], INDEX = SUMM[,1], FUN = mean);
Mf_nonupt    <- tapply(X = SUMM[,18], INDEX = SUMM[,1], FUN = mean);

>>>>

---
title: "A general model for the evolution of nuptial gift-giving"
author: |
  | Author list (anonymised)
date: '[1] Author institutions and email'
output:
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - evolution-letters.csl
    reference_docx: docx_template.docx
  html_document: default
  pdf_document:
    fig_caption: yes
    keep_tex: yes
bibliography: nuptial_refs.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{natbib}
- \usepackage{lineno}
- \usepackage{caption}
- \usepackage[utf8]{inputenc}
- \bibliographystyle{amnatnat}
- \linenumbers
- \modulolinenumbers[2]
linestretch: 2
link-citations: yes
linkcolor: blue
csl: evolution-letters.csl
biblio-style: apalike
---

@@

We ran simulations in which male search time and female choosiness evolved from an ancestral state of no searching and no choosiness. 
In all simulations, $N$ was initialised at 1000 and $K = 1000$. 
Simulations ran for $t_{max} = 10000$ time steps. 
We set $T_{\mathrm{f}} = 2$, $\psi = 3$, and $\lambda = 1$ for all simulations, and we simulated across a range of $\alpha = \{0.1, 0.2, ..., 1.9, 2.0\}$ and $\gamma = \{0, 0.1, ..., 1.9, 2.0\}$ parameter values for 5000 replicates.
In a sensitivity analysis, we also re-ran simulations for values of $T_{\mathrm{f}} = 10$ and $\psi = \{1, 2, 4, 6 \}$ (see Supporting Information S4), and simulated the evolution of male nuptial gift searching and female choosiness (fixing $T_{\mathrm{m}} = \alpha$) separately (see Supporting Information S4).

We additionally ran simulations in which males adopted one of two search strategies: either search in time-out until a gift is obtained, or do not search at all.
This simulations were therefore more closely aligned with the assumptions of our analytical model, but results did not differ qualitatively (Supporting Information S5).
The strategy adopted by a male in these simulations was based on an underlying trait for the probability of nuptial gift search.
Like $\phi^{i}$, this trait was underpinned by a diploid genotype in which alleles had additive effects.

Summary statistics for mean trait values, population size, sex ratios, proportion of females and males in and out of the mating pool, and mean number of encounters per female and male within the mating pool were all calculated in the last time step. 
The C code used for simulating these IBMs also allows for the reporting of summary statistics in each time step. 
Additionally, it can simulate explicit space and individual movement through the landscape. 
A neutral evolving trait was also modelled to ensure that the code functioned as intended, and processes were compartmentalised into individual functions to facilitate code testing. 
All code is publicly available on GitHub (https://github.com/AUTHOR_NAME/REPOSITORY_NAME).

@@

For males, if nuptial gifts are not abundant and thus require a long time to find (i.e., high $\alpha$), or if males encounter many females per unit time (i.e., high $R / (1+\beta)$), then the nuptial gift must result in a high fitness increment for selection to favour searching. 
For females, if $\gamma$ is sufficiently high, fitness is increased by rejecting males without gifts and mating only with males that provide nuptial gifts. 
As offspring processing time ($T_{\mathrm{f}}$), mate encounter rate ($R\beta / (\beta + 1)$), or the probability of a male finding a nuptial gift ($1 - \exp(-T_{\mathrm{m}}/\alpha)$ ) decrease, the fecundity increment above which selection will favour choosiness ($\gamma_{\mathrm{f}}$) increases.
This can be understood intuitively by realising that rejecting a prospective male represents an opportunity cost for the female.

```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.cap = "Coevolution of male search and female choosiness as a function of nuptial gift search time ($\\alpha$). Points show where the lower 95% confidence interval of female choosiness (red) and male search (blue) exceeds zero, indicating evolution of choosiness or nuptial gift search. Each point includes data from 5000 replicate simulations with identical starting conditions. Zones are identified that correspond to areas where males search and females are choosy (D), males search but females are not choosy (B), and males do not search and females are not choosy (A), as also depicted in Figure 2. The number of individuals in the population remained at or near carrying capacity of $K = 1000$. In each time step, up to 3000 total pair-wise interactions occurred. Expected female processing time was set to $T_{\\mathrm{f}}=2$ time steps, and $\\gamma$ and $\\alpha$ values in the range [0.0, 2.0] and [0.1, 2.0], respectively, were used."}
SUMM         <- read.csv(file = "inst/summary_coevolving.csv");
alpha_vals   <- sort(unique(SUMM[,1]));  
TM_vals      <- tapply(X = SUMM[,3],  INDEX = SUMM[,1], FUN = mean);
m_vals       <- tapply(X = SUMM[,13], INDEX = SUMM[,1], FUN = mean);
beta_vals    <- tapply(X = SUMM[,14], INDEX = SUMM[,1], FUN = mean);
Male_M       <- tapply(X = SUMM[,15], INDEX = SUMM[,1], FUN = mean);
Female_M     <- tapply(X = SUMM[,16], INDEX = SUMM[,1], FUN = mean);
Mf_nupt      <- tapply(X = SUMM[,17], INDEX = SUMM[,1], FUN = mean);
Mf_nonupt    <- tapply(X = SUMM[,18], INDEX = SUMM[,1], FUN = mean);

### 2024-12-20 ms.tex
Update last comments based on round of revisions and Evol Lett comments

We ran simulations in which male search time and female choosiness
evolved from an ancestral state of no searching and no choosiness. In
all simulations, \(N\) was initialised at 1000 and \(K = 1000\).
Simulations ran for \(t_{max} = 10000\) time steps. We set
\(T_{\mathrm{f}} = 2\), \(\psi = 3\), and \(\lambda = 1\) for all
simulations, and we simulated across a range of
\(\alpha = \{0.1, 0.2, ..., 1.9, 2.0\}\) and
\(\gamma = \{0, 0.1, ..., 1.9, 2.0\}\) parameter values for 5000
replicates. In a sensitivity analysis, we also re-ran simulations for
values of \(T_{\mathrm{f}} = 10\) and \(\psi = \{1, 2, 4, 6 \}\) (see
Supporting Information S3), and simulated the evolution of male nuptial
gift searching and female choosiness (fixing
\(T_{\mathrm{m}} = \alpha\)) separately (see Supporting Information S4).
Summary statistics for mean trait values, population size, sex ratios,
proportion of females and males in and out of the mating pool, and mean
number of encounters per female and male within the mating pool were all
calculated in the last time step. The C code used for simulating these
IBMs also allows for the reporting of summary statistics in each time
step. Additionally, it can simulate explicit space and individual
movement through the landscape. A neutral evolving trait was also
modelled to ensure that the code functioned as intended, and processes
were compartmentalised into individual functions to facilitate code
testing. All code is publicly available on GitHub
(\url{https://github.com/AUTHOR_NAME/REPOSITORY_NAME}).

@@

\begin{figure}
\centering
\includegraphics{ms_files/figure-latex/unnamed-chunk-3-1.pdf}
\caption{Fitness thresholds above which males increase their fitness by
searching for nuptial gifts (blue lines; Eq. 2) and females increase
their fitness by rejecting males that do not offer gifts (red lines; Eq.
3). Parameter space includes areas in which males do not search for
nuptial gifts and females are not choosy (A), males search but females
are not choosy (B), females would be choosy but males do not search (C),
and males search and females are choosy (D). Arrows in panel a indicate
the effect of increasing encounter rate (\(R\)), female time-out
(\(T_{\mathrm{f}}\)), and male search time (\(T_{\mathrm{m}}\)). As an
example, trajectories for \(T_{\mathrm{f}} = 2\), and
\(T_{\mathrm{m}} = \alpha\) are shown for values of \(R=2\) (panel a)
and \(R = 1\) (panel b). Females are assumed to be the choosy sex, which
is maintained as long as \(\alpha < T_{\mathrm{f}}\).}
\end{figure}

For males, if nuptial gifts are not abundant and thus require a long
time to find (i.e., high \(\alpha\)), or if males encounter many females
per unit time (i.e., high \(R / (1+\beta)\)), then the nuptial gift must
result in a high fitness increment for selection to favour searching.
For females, if \(\gamma\) is sufficiently high, fitness is increased by
rejecting males without gifts and mating only with males that provide
nuptial gifts. As offspring processing time (\(T_{\mathrm{f}}\)), mate
encounter rate (\(R\beta / (\beta + 1)\)), or the probability of a male
finding a nuptial gift (\(1 - \exp(-T_{\mathrm{m}}/\alpha)\) ) decrease,
the threshold value of fitness above which selection will favour
choosiness (\(\gamma_{\mathrm{f}}\)) increases. This can be understood
intuitively by realising that rejecting a prospective male represents an
opportunity cost for the female.

\begin{figure}
\centering
\includegraphics{ms_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{The coevolution of male search and female choosiness as a
function of nuptial gift search time (\(\alpha\)). Points show where the
lower 95\% confidence interval of female choosiness (red) and male
search (blue) exceeds zero, indicating evolution of choosiness or
nuptial gift search. Each point includes data from 5000 replicate
simulations with identical starting conditions. Zones are identified
that correspond to areas where males search and females are choosy (D),
males search but females are not choosy (B), and males do not search and
females are not choosy (A), as also depicted in Figure 2. The number of
individuals in the population remained at or near carrying capacity of
\(K = 1000\). In each time step, up to 3000 total pair-wise interactions
occurred. Expected female processing time was set to
\(T_{\mathrm{f}}=2\) time steps, and \(\gamma\) and \(\alpha\) values in
the range {[}0.0, 2.0{]} and {[}0.1, 2.0{]}, respectively, were used.}
\end{figure}

>>>>

We ran simulations in which male search time and female choosiness
evolved from an ancestral state of no searching and no choosiness. In
all simulations, \(N\) was initialised at 1000 and \(K = 1000\).
Simulations ran for \(t_{max} = 10000\) time steps. We set
\(T_{\mathrm{f}} = 2\), \(\psi = 3\), and \(\lambda = 1\) for all
simulations, and we simulated across a range of
\(\alpha = \{0.1, 0.2, ..., 1.9, 2.0\}\) and
\(\gamma = \{0, 0.1, ..., 1.9, 2.0\}\) parameter values for 5000
replicates. In a sensitivity analysis, we also re-ran simulations for
values of \(T_{\mathrm{f}} = 10\) and \(\psi = \{1, 2, 4, 6 \}\) (see
Supporting Information S4), and simulated the evolution of male nuptial
gift searching and female choosiness (fixing
\(T_{\mathrm{m}} = \alpha\)) separately (see Supporting Information S4).

We additionally ran simulations in which males adopted one of two search
strategies: either search in time-out until a gift is obtained, or do
not search at all. This simulations were therefore more closely aligned
with the assumptions of our analytical model, but results did not differ
qualitatively (Supporting Information S5). The strategy adopted by a
male in these simulations was based on an underlying trait for the
probability of nuptial gift search. Like \(\phi^{i}\), this trait was
underpinned by a diploid genotype in which alleles had additive effects.

Summary statistics for mean trait values, population size, sex ratios,
proportion of females and males in and out of the mating pool, and mean
number of encounters per female and male within the mating pool were all
calculated in the last time step. The C code used for simulating these
IBMs also allows for the reporting of summary statistics in each time
step. Additionally, it can simulate explicit space and individual
movement through the landscape. A neutral evolving trait was also
modelled to ensure that the code functioned as intended, and processes
were compartmentalised into individual functions to facilitate code
testing. All code is publicly available on GitHub
(\url{https://github.com/AUTHOR_NAME/REPOSITORY_NAME}).

@@

\begin{figure}
\centering
\includegraphics{ms_files/figure-latex/unnamed-chunk-3-1.pdf}
\caption{Fecundity increment above which males increase their fitness by
searching for nuptial gifts (blue lines; Eq. 2) and females increase
their fitness by rejecting males that do not offer gifts (red lines; Eq.
3). Parameter space includes areas in which males do not search for
nuptial gifts and females are not choosy (A), males search but females
are not choosy (B), females would be choosy but males do not search (C),
and males search and females are choosy (D). Arrows in panel a indicate
the effect of increasing encounter rate (\(R\)), female time-out
(\(T_{\mathrm{f}}\)), and male search time (\(T_{\mathrm{m}}\)). As an
example, curves for \(T_{\mathrm{f}} = 2\), and
\(T_{\mathrm{m}} = \alpha\) are shown for values of \(R=2\) (panel a)
and \(R = 1\) (panel b). Females are assumed to be the choosy sex, which
is maintained as long as \(\alpha < T_{\mathrm{f}}\).}
\end{figure}

For males, if nuptial gifts are not abundant and thus require a long
time to find (i.e., high \(\alpha\)), or if males encounter many females
per unit time (i.e., high \(R / (1+\beta)\)), then the nuptial gift must
result in a high fitness increment for selection to favour searching.
For females, if \(\gamma\) is sufficiently high, fitness is increased by
rejecting males without gifts and mating only with males that provide
nuptial gifts. As offspring processing time (\(T_{\mathrm{f}}\)), mate
encounter rate (\(R\beta / (\beta + 1)\)), or the probability of a male
finding a nuptial gift (\(1 - \exp(-T_{\mathrm{m}}/\alpha)\) ) decrease,
the fecundity increment above which selection will favour choosiness
(\(\gamma_{\mathrm{f}}\)) increases. This can be understood intuitively
by realising that rejecting a prospective male represents an opportunity
cost for the female.

\begin{figure}
\centering
\includegraphics{ms_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{Coevolution of male search and female choosiness as a function
of nuptial gift search time (\(\alpha\)). Points show where the lower
95\% confidence interval of female choosiness (red) and male search
(blue) exceeds zero, indicating evolution of choosiness or nuptial gift
search. Each point includes data from 5000 replicate simulations with
identical starting conditions. Zones are identified that correspond to
areas where males search and females are choosy (D), males search but
females are not choosy (B), and males do not search and females are not
choosy (A), as also depicted in Figure 2. The number of individuals in
the population remained at or near carrying capacity of \(K = 1000\). In
each time step, up to 3000 total pair-wise interactions occurred.
Expected female processing time was set to \(T_{\mathrm{f}}=2\) time
steps, and \(\gamma\) and \(\alpha\) values in the range {[}0.0, 2.0{]}
and {[}0.1, 2.0{]}, respectively, were used.}
\end{figure}
